
---
title: "Week 12: Animal movement analysis"
output: github_document
---

```{r global_options, eval=T, echo=F,results='hide', error=F, warning=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',  warning=FALSE, message=FALSE, results='hide')
rm(list=ls())
library(adehabitatLT)
```

# What is a trajectory

## Type I, no time information recorded
## Type II: regular and irregular trajectories

# Trajectory representation

A trajectory are composed by a list of bursts, which can be characterized by:
x, y, t, dx, dy, dist, dt, R2n, abs.angel, rel.angel

The basic analysis element is 'ltraj' and it needs coordinates and the time stamp to construct a trajectory. The 'ltraj' precompute descriptive parameters of the trajectory including the time and distance lags between the relocations.

```{r trj1, eval=T, echo=T}
data(whale)
plot(whale)
class(whale)

# The 'ltraj' can be constructed from a data frame

data(puechabonsp)
locs <- puechabonsp$relocs
locs <- as.data.frame(locs)
head(locs)

class(locs$Date)

da <- as.POSIXct(strptime(as.character(locs$Date),"%y%m%d"))

puech <- as.ltraj(xy = locs[,c("X","Y")], date = da, id = locs$Name)
puech

head(puech[[1]])
plot(puech)

# The trajeocty can be regular or irregular observed

is.regular(puech)

# plotltr can help investigate the observation intervals

plotltr(puech, "dt/3600/24")

# divide the trajectory into bursts if the interval is longer than a threshold value

foo <- function(dt) {
     return(dt> (100*3600*24))
 }

puech2 <- cutltraj(puech, "foo(dt)", nextr = TRUE)
puech2

burst(puech2)[3:4] <- c("Chou.1992", "Chou.1993")
puech2


# The function which.ltraj can also be used to identify the bursts satisfying a condition. For example, imagine that we want to identify the bursts where the distance between successive relocations was greater than 2000 metres at least once:

bu <- which.ltraj(puech2, "dist>2000")

#This data frame contains the ID, burst ID and relocation numbers satisfying the specified criterion. We can then extract the bursts satisfying this criterion:

(puech2[burst(puech2)%in%bu$burst])

plot(puech2[5])


# make the irregular observed trajectory into regular ones by filling in missing values

plotltr(puech2, "dt/3600/24")

refda <- strptime("00:00", "%H:%M")
refda

puech3 <- setNA(puech2, refda, 1, units = "day")
puech3

plotltr(puech3)

# rounding the time to define a regular trajectory

data(ibexraw)

plotltr(ibexraw, "dt/3600")

refda <- strptime("2003-06-01 00:00", "%Y-%m-%d %H:%M", tz="Europe/Paris")
ib2 <- setNA(ibexraw, refda, 4, units = "hour")
ib2

plotltr(ib2, "dt/3600")

ib3 <- sett0(ib2, refda, 4, units = "hour")
ib3
plotltr(ib3, "dt/3600")


## align the trajectory with the same duration

is.sd(ib3)

ib4 <- set.limits(ib3, begin = "2003-06-01 00:00", dur = 14, units = "day", pattern = "%Y-%m-%d %H:%M", tz="Europe/Paris")

ib4

is.sd(ib4)

di <- sd2df(ib4, "dist")

```

# Analysis of trajectory


```{r trj2, eval=T, echo=T}

## randomness of the mising values

runsNAltraj(ib4)

# This trajectory is regular. The bear was monitored during one month, with one relocation every 30 minutes.

data(bear)
runsNAltraj(bear)
plotNAltraj(bear)


# convert the time stamped trajectory into non time-stampled ones
bearI <- typeII2typeI(bear)
plot(bearI)


## rediscretize this trajectory with constant step length of 500 metres:
bearIr <- redisltraj(bearI, 500)
bearIr

plot(bearIr)

# the geometrical properties can be studied by examining the distribution of the relative angles.

sliwinltr(bearIr, function(x) mean(cos(x$rel.angle)), type="locs", step=30)

cosrelangle <- cos(bearIr[[1]]$rel.angle)

head(cosrelangle)


## interpolate for by equal time intervals

data(porpoise)
porpoise2 <- redisltraj(na.omit(porpoise[1:3]), 86400, type="time")
plot(porpoise2[2])

trajdyn(ib4)

```
# Autocorrelation analysis of trajectory


```{r trj3, eval=T, echo=T}
## test autocorrelation of the parameters

wawotest(bear)

plotltr(bear, "dist")

sliwinltr(bear, function(x) mean(na.omit(x$dist)), 5*48, type="locs")

# ACF function for distance

acfdist.ltraj(bear, lag=5, which="dist")

# Testing the relative angels
testang.ltraj(bear, "relative")

acfang.ltraj(bear, lag=5)
```

# Segment of a spatial trajectory

```{r trj4, eval=T, echo=T}

data(porpoise)
gus <- porpoise[1]
gus
plot(gus)

acfdist.ltraj(gus, "dist", lag=20)
plotltr(gus, "dist")


# Segmentation method by Gueguen (2001)

(tested.means <- round(seq(0, 130000, length = 10), 0))
(limod <- as.list(paste("dnorm(dist, mean =", tested.means, ", sd = 5000)")))
mod <- modpartltraj(gus, limod)

mod

bestpartmod(mod)

(pm <- partmod.ltraj(gus, 4, mod))

plot(pm)

plotltr(gus, "dist")

tmp <- lapply(1:length(pm$ltraj), function(i){coul <- c("red","green","blue")[as.numeric(factor(pm$stats$mod))[i]]
    lines(pm$ltraj[[i]]$date, rep(tested.means[pm$stats$mod[i]], nrow(pm$ltraj[[i]])),col=coul, lwd=2)})


```

# Models of animal movements

Brownian motion, random walk, Brownian bridge motion (i.e. a Brownian motion
constrained by a fixed start and end point), Levy flight

```{r trj5, eval=T, echo=T}

# Simulate a null model of trajectory

sim <- simm.crw(1:1000, r=0.95)
plot(sim, addp=F)

sim <- simm.levy(x=1:1000)
plot(sim, addp=F)


# Null model
# A commonly used method in community ecology.
#  Gotelli and Graves (1996) defined the null model as “a pattern generating model that is based on randomization of ecological data or random sampling from a known or imag- ined distribution. The null model is designed with respect to some ecological or evolutionary process of interest. Certain elements of the data are held constant, and others are allowed to vary stochastically to create new assemblage patterns. The randomization is designed to produce a pattern that would be expected in the absence of a particular ecological mechanism”.


data(puechcirc)
data(puechabonsp)
xo <- coordinates(puechabonsp$map)
boar1 <- puechcirc[1]
plot(boar1, spixdf=puechabonsp$map, xlim=range(xo[,1]), ylim=range(xo[,2]))

plotfun <- function(x, par) {
 image(par)
     points(x[,1:2], pch=16)
     lines(x[,1:2])
 return(x) }


confun <- function(x, par) {
     ## Define a SpatialPointsDataFrame from the trajectory
     coordinates(x) <- x[,1:2]
     ## overlap the relocations x to the elevation map par
     jo <- join(x, par)
     ## checks that there are no missing value
     res <- all(!is.na(jo))
     ## return this check
     return(res) }

 nmo2 <- NMs.randomShiftRotation(na.omit(boar1), rshift = TRUE, rrot = TRUE,
                                 rx = range(xo[,1]), ry = range(xo[,2]),
                                 treatment.func = plotfun,
                                 treatment.par = puechabonsp$map[,1],
                                 constraint.func = confun,
                                 constraint.par = puechabonsp$map[,1], nrep=9)

set.seed(90909)
par(mfrow=c(3,3), mar=c(0,0,0,0))
resu <- testNM(nmo2, count = FALSE)

# compare the variance of the elevation.

varel <- function(x, par) {
     coordinates(x) <- x[,1:2]
     jo <- join(x, par)
     return(var(jo)) }

nmo3 <- NMs.randomShiftRotation(na.omit(boar1), rshift = TRUE, rrot = TRUE,
                                 rx = range(xo[,1]), ry = range(xo[,2]),
                                 treatment.func = varel,
                                 treatment.par = puechabonsp$map[,1],
                                 constraint.func = confun,
                                 constraint.par = puechabonsp$map[,1], nrep=99)

sim <- testNM(nmo3, count=FALSE)

(obs <- varel(na.omit(boar1)[[1]], puechabonsp$map[,1]))
(ran <- as.randtest(unlist(sim[[1]]), obs))
plot(ran)


## Null model with multiple animals

data(puechcirc)
plot(puechcirc)


(boar <- bindltraj(puechcirc))

nmo4 <- NMs.randomShiftRotation(na.omit(boar), rshift = TRUE, rrot = TRUE,
                                 rx = range(xo[,1]), ry = range(xo[,2]),
                                 treatment.func = varel,
                                 treatment.par = puechabonsp$map[,1],
                                 constraint.func = confun,
                                 constraint.par = puechabonsp$map[,1], nrep=99)
sim2 <- testNM(nmo4, count=FALSE)

(obs <- lapply(na.omit(boar), function(x) {varel(x, puechabonsp$map[,1])}))

lapply(1:2, function(i) {as.randtest(unlist(sim2[[i]]), obs[[i]])})

```
